---
title: "combine_am_iucn_results"
output: html_document
---

## Summary

Here I combine the results from the separate IUCN and AM analyses, for species- countries and species - cells, as well as the sensitivity analyses 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(data.table)

```

## SPECIES

These results have species information and the number of countries they occur in (using 10 cell cutoff)

```{r combine spp results}

## load results

am_spp_summary<- fread(here("_data/am_spp_summary_10cells.csv"))

iucn_spp_summary <- fread(here("_data/iucn_spp_summary_10cells.csv"))

names(am_spp_summary)
names(iucn_spp_summary)

```

# Combine species results

```{r combine spp results}

## some checks

nrow(am_spp_summary) # n = 18353
nrow(iucn_spp_summary) # n = 9900

unique(am_spp_summary$kingdom)
unique(iucn_spp_summary$kingdom) # no chromista or fungi 

# organize and rename cols to bind
str(am_spp_summary)
am_spp_summary$category <- "None"
am_spp_summary <- am_spp_summary %>% rename(num_eez = num_countries)

names(iucn_spp_summary)
iucn_spp_summary <- iucn_spp_summary %>% rename(species_id = iucn_sid)
iucn_spp_summary$species_id <- as.character(iucn_spp_summary$species_id)

combined_spp_summary <- bind_rows(am_spp_summary, iucn_spp_summary)
nrow(combined_spp_summary) # n = 28253 (with cane toad still in there)

### xbd spread

table(combined_spp_summary$xbd)
# singlecountry transboundary 
  #     2691         25562

head(combined_spp_summary)

# Change kingdoms and phyla names to all caps
combined_spp_summary<- combined_spp_summary %>% 
  mutate(kingdom = toupper(kingdom), phylum = toupper(phylum), class = toupper(class))
unique(combined_spp_summary$kingdom)
unique(combined_spp_summary$phylum) 
combined_spp_summary %>% arrange(class) %>% distinct(class) # 70 classes
# NB difference in some class classificaiton between iucn and AM (e.g. elasmobranchii v chondricthyes)

```

## Wrangle taxon groupings and threat categories

Using the 10 cell cutoff version. Protozoa, Bacteria, and Chromista were removed

```{r taxonomic breakdown by country}

## Make new taxa groups

combined_spp_summary %>% distinct(kingdom, phylum, class) %>% arrange(kingdom)
combined_spp_summary %>% distinct(kingdom, phylum, class) %>% arrange(class)
  # 7 phyla of worms and sea slugs not assigned classes

## identify vertebrates 
combined_spp_summary %>% distinct(kingdom, phylum, class) %>% arrange(phylum)
  # chordates but not verts: 
    ## ASCIDIACEA (tunicates), THALIACEA (other tunicates), APPENDICULARIA (more tunicates), CEPHALOCHORDATA (lancelets)

verts <- c("ACTINOPTERYGII","AMPHIBIA","AVES","CEPHALASPIDOMORPHI","CHONDRICHTHYES","ELASMOBRANCHII","HOLOCEPHALI","MAMMALIA","MYXINI","REPTILIA","SARCOPTERYGII") 
# NB CEPHALASPIDOMORPHI are sometimes used for lampreys but others consider them an extinct class

combined_spp_summary <- combined_spp_summary %>%
  mutate(tax_group = case_when(
    kingdom == "PLANTAE" ~ "Plant",
    kingdom == "ANIMALIA" & class %in% verts ~ "Vertebrate",
    TRUE ~ "Invertebrate"))

table(combined_spp_summary$tax_group)
# Invertebrate        Plant   Vertebrate 
#     13239          316        14698 

#<><><><><
## change redlist cateogries 

unique(combined_spp_summary$category)

combined_spp_summary %>% filter(category %in% c("LR/lc","LR/nt")) %>% distinct(sciname, class)
# only 5 species, so add to LC category
combined_spp_summary$category <- combined_spp_summary$category %>% fct_collapse(LC = c("LC","LR/lc","LR/nt"))

fwrite(combined_spp_summary, "_data/combined_spp_summary.csv")

```

## COUNTRIES

These results have each country and the number of single country (sc) and tb species they have

```{r combine country results}

## Load data (10 cell cutoff results)

iucn_eez_summary <- fread(here("_data/iucn_eez_summary_10cells.csv"))
names(iucn_eez_summary)

am_eez_summary <- fread(here("_data/am_eez_summary_10cells.csv"))
names(am_eez_summary)
```

# Combine countries results

```{r combine countries reults}

#<><<> 10 cells

names(am_eez_summary)
names(iucn_eez_summary)

# change AM col names for bind
am_eez_summary <- am_eez_summary %>% rename(num_spp_tb = num_am_spp_tb, num_spp_sc = num_am_spp_sc)

## join

combined_eez_summary <- bind_rows(iucn_eez_summary, am_eez_summary) %>%
  filter(!is.na(rgn_id_2)) %>%
  replace_na(list(num_spp_tb = 0, num_spp_sc = 0)) %>%
  group_by(rgn_id_2) %>%
  summarise(num_spp_tb = sum(num_spp_tb),
            num_spp_sc = sum(num_spp_sc))

n_distinct(combined_eez_summary$rgn_id_2) # n = 228
nrow(combined_eez_summary) # n = 228

combined_eez_summary %>% select_if(function(x) any(is.na(x))) # none

#<><><>
## add region names

rgn_lookup <- read_csv(here("_data/rgn_lookup.csv"))

anti_join(rgn_lookup, combined_eez_summary, by = "rgn_id_2")
#  rgn_id_2    rgn_nam_2
#      188     Slovenia
#      185       Monaco
#       60    Gibraltar
#      192         Iraq
#      215       Jordan
#      220 Sint Maarten

combined_eez_summary <- left_join(combined_eez_summary, lookup, by = "rgn_id_2") %>% filter(!is.na(rgn_id_2))

n_distinct(combined_eez_summary$rgn_id_2) # n = 228

##<><<><
# Add thraetened species counts

eez_summary_10cells_thr <- fread(here("_data/eez_summary_10cells_thr.csv"))

## Add th spp counts to combined summary
  
combined_eez_summary <- left_join(combined_eez_summary, eez_summary_10cells_thr, by="rgn_id_2") %>%
    replace_na(list(num_spp_tb = 0, num_spp_sc = 0, num_spp_tb_th = 0, num_spp_sc_th= 0)) 
names(combined_eez_summary)

#<><><>><
## quick explore

combined_eez_summary %>% arrange(desc(num_spp_sc))
combined_eez_summary %>% arrange(desc(num_spp_tb))
combined_eez_summary %>% arrange(desc(num_spp_sc_th))
combined_eez_summary %>% arrange(desc(num_spp_tb_th))

## save 
fwrite(combined_eez_summary, "_data/combined_eez_summary.csv")


```

## Sensitivity analyses

5 percent cutoff scenario

```{5 percent cutoff results}

##<><>><<> 5% cutoff species summary

iucn_spp_5perc <- fread(here("_data/cutoff_scenarios/iucn_spp_summary_5perc.csv"))

am_spp_5perc <- fread(here("_data/cutoff_scenarios/am_spp_summary_5perc.csv"))

# organize and rename cols to bind
names(am_spp_5perc)
nrow(am_spp_5perc) # n = 22410

names(iucn_spp_5perc)
nrow(iucn_spp_5perc) # n = 10340
str(iucn_spp_5perc) # here the species id is just the number

iucn_spp_5perc <- iucn_spp_5perc %>% rename(species_id = iucn_sid, cell_count_cutoff = cell_cutoff)

iucn_spp_5perc$species_id <- as.character(iucn_spp_5perc$species_id)

spp_summary_5perc <- bind_rows(am_spp_5perc, iucn_spp_5perc)
nrow(spp_summary_5perc) # n = 32750

# Change kingdoms and phyla names to all caps
spp_summary_5perc <- spp_summary_5perc %>% 
  mutate(kingdom = toupper(kingdom), phylum = toupper(phylum))
unique(spp_summary_5perc$kingdom)
unique(spp_summary_5perc$phylum)

table(spp_summary_5perc$xbd)
# singlecountry transboundary 
 #        3201         29549


#<><<> 5 percent cutoff Countries summary

## load data

iucn_countries_5perc <- fread(here("_data/cutoff_scenarios/iucn_country_summary_5perc.csv"))

am_countries_5perc <- fread(here("_data/cutoff_scenarios/am_country_summary_5perc.csv"))

## organize and rename cols to bind

names(iucn_countries_5perc)
iucn_countries_5perc <- iucn_countries_5perc %>% rename(iucn_num_sc = num_spp_sc, iucn_num_tb = num_spp_tb)

names(am_countries_5perc)
am_countries_5perc <- am_countries_5perc %>% rename(am_num_sc = num_spp_sc, am_num_tb = num_spp_tb)

countries_summary_5perc <- full_join(iucn_countries_5perc, am_countries_5perc, by = "rgn_name", fill = 0)
n_distinct(countries_summary_5perc$rgn_name) # n = 218
nrow(countries_summary_5perc) 

countries_summary_10cells %>% select_if(function(x) any(is.na(x))) # lots of NAs

countries_summary_5perc <- countries_summary_5perc%>% 
  replace_na(list(iucn_num_sc = 0, iucn_num_tb = 0, am_num_tb = 0, am_num_sc = 0))

countries_summary_5perc <- countries_summary_5perc %>% rowwise() %>%
  mutate(num_spp_sc = sum(iucn_num_sc,am_num_sc), num_spp_tb = sum(iucn_num_tb,am_num_tb)) %>%
  select(-c(2:5))

head(countries_summary_5perc) 
countries_summary_5perc %>% arrange(desc(num_spp_sc))
countries_summary_5perc %>% arrange(desc(num_spp_tb))

```

## compare to no cutoff for occurrence

combine iucn and am results for A) species and B) countries without any cutoff for occurring in an EEZ

```{r }

#<><><><><< SPECIES SUMMARY

iucn_spp <- fread(here("_data/cutoff_scenarios/iucn_spp_summary_nocutoff.csv"))
am_spp <-  fread(here("_data/cutoff_scenarios/am_spp_summary_nocutoff.csv"))
  
## some checks
nrow(am_spp) # n = 22410
unique(am_spp$kingdom) 

## organize and rename cols to bind
names(am_spp)
str(am_spp_10cells)
am_spp$category <- "NA_AM"

names(iucn_spp)
unique(iucn_spp$kingdom) # no chromista or fungi so they must not have maps for them
nrow(iucn_spp) # n = 10340
str(iucn_spp) # here the species id is just the number and is numeric
iucn_spp <- iucn_spp %>% rename(species_id = iucn_sid)
iucn_spp$species_id <- as.character(iucn_spp$species_id)

spp_summary_nocutoff <- bind_rows(am_spp, iucn_spp)
nrow(spp_summary_nocutoff) # 32750

table(spp_summary_nocutoff$xbd)
# singlecountry transboundary 
 #        2762         29988
2762/(2762+29988)*100 # 8.43% sc

## Fix different formats of factor levels>> Change kingdoms and phyla names to all caps
spp_summary_nocutoff <- spp_summary_nocutoff %>% 
  mutate(kingdom = toupper(kingdom), phylum = toupper(phylum))
unique(spp_summary_nocutoff$kingdom)
unique(spp_summary_nocutoff$phylum)

table(spp_summary_nocutoff$xbd)
# singlecountry transboundary 
        # 2762         29988 
         
         
#<><><>><<> COUNTRIES SUMMARY

iucn_countries <- fread(here("_data/cutoff_scenarios/iucn_country_summary_nocutoff.csv"))
am_countries <- fread(here("_data/cutoff_scenarios/am_country_summary_nocutoff.csv"))

names(iucn_countries)
iucn_countries <- iucn_countries %>% rename(iucn_num_sc = num_spp_sc, iucn_num_tb = num_spp_tb)

names(am_countries)
am_countries <- am_countries %>% rename(am_num_sc = num_spp_sc, am_num_tb = num_spp_tb)

countries_summary_nocutoff <- full_join(iucn_countries, am_countries, by = "rgn_name", fill = 0)
n_distinct(countries_summary_nocutoff$rgn_name) # n = 222
nrow(countries_summary_nocutoff) # n = 222 >> with the cutoffs there were fewer regions

countries_summary_nocutoff %>% select_if(function(x) any(is.na(x))) # lots of NAs

countries_summary_nocutoff <- countries_summary_nocutoff %>% 
  replace_na(list(iucn_num_sc = 0, iucn_num_tb = 0, am_num_tb = 0, am_num_sc = 0))

countries_summary_nocutoff <- countries_summary_nocutoff %>% rowwise() %>%
  mutate(num_spp_sc = sum(iucn_num_sc,am_num_sc), num_spp_tb = sum(iucn_num_tb,am_num_tb)) %>%
  select(-c(2:5))

head(countries_summary_nocutoff, 20)
countries_summary_nocutoff %>% arrange(desc(num_spp_sc))
countries_summary_nocutoff %>% arrange(desc(num_spp_tb))

```

